---
title: Data types and structures
author: Marie-Hélène Burle
---

:::{.def}

It might be time to talk a bit more formally about the various data types and structures available in R. The goal of this course is not to get bogged down in the nitty-gritty of R syntax, so this section is kept very short.

:::

## Data types

```{r}
typeof("Some words")
typeof(2)
typeof(2.0)
typeof(2L)
typeof(TRUE)
```

## Data structures

| Dimension | Homogeneous   | Heterogeneous  |
|-----------|---------------|----------------|
| 1 d       | Atomic vector | List           |
| 2 d       | Matrix        | Data frame     |
| 3 d       | Array         |                |

### Atomic vectors

```{r}
c(2, 4, 1)
str(c(2, 4, 1))

c(2.2, 4.4, 1.0)
str(c(2.2, 4.4, 1.0))

1:3
str(1:3)

c("some", "random", "words")
str(c("some", "random", "words"))
```

### Matrices

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m
str(m)
```

### Arrays

```{r}
a <- array(as.double(1:24), c(3, 2, 4))
a
str(a)
```

### Lists

```{r}
l <- list(2L, 3, c(2, 1), FALSE, "string")
l
str(l)
```

### Data frames

```{r}
d <- data.frame(
  country = c("Canada", "USA", "Mexico"),
  var = c(2.9, 3.1, 4.5)
)
d
str(d)
```
