---
title: Data exploration
author: Marie-Hélène Burle
---

:::{.def}

An important first step of data analysis is to have a look at the data. In this section, we will explore the `us_contagious_diseases` dataset from the [dslabs package](https://cran.r-project.org/web/packages/dslabs/index.html).

:::

## Load the dslabs package

This package contains a number of datasets. To access any of them, we first need to load the package:

```{r}
library(dslabs)
```

:::{.info}

`library()` is a *function*:

```{r}
class(library)
```

Functions are the "verbs" of programming languages. They do things.

`library()` is a function that loads packages into the current session so that their content becomes available.

`dslabs` is the *argument* that we pass to the function `library()`: it is this particular packages that we are loading in the session here.

`class()` is also a function: it tells what class an object belongs to. In `class(library)`, `library` is the argument of the function `class()`.

:::

## Basic exploration

```{r}
head(us_contagious_diseases)
head(us_contagious_diseases, 15)
tail(us_contagious_diseases)
names(us_contagious_diseases)
str(us_contagious_diseases)
typeof(us_contagious_diseases)
class(us_contagious_diseases)
```

```{.r}
View(us_contagious_diseases)
```

```{r}
dim(us_contagious_diseases)
ncol(us_contagious_diseases)
nrow(us_contagious_diseases)
```

```{r}
length(us_contagious_diseases)
length(us_contagious_diseases$disease)
```

## Indexing

```{r}
us_contagious_diseases[1, 1]
us_contagious_diseases[1, 2]
```

```{r}
us_contagious_diseases[1, ]
us_contagious_diseases[3000, ]
```

## Summary statistics

```{r}
summary(us_contagious_diseases)
```
